---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyGWAS

<!-- badges: start -->
<!-- badges: end -->

The goal of tidyGWAS is to provide an easy-to use R package written in the typical 
tidyverse/dplyr syntax to make cleaning and collection of sumstats simple, and to
stop analysts for falling to common error. In addition we provide helpers to standardise
GWAS data: Reparation of stats columns where possible (missing B/SE/P), as well
as by default providing coordinates on both GRCh37 and GRCh38.

An early design goal of tidyGWAS was to keep as many rows as possible. Our default options 
are therefore to keep multi-allelic SNPs or ambigious SNPs, which we suggest to handle
depending on what downstream analysis tool you wish to use.






## Installation

You can install the development version of tidyGWAS from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Ararder/tidyGWAS")

# To install the Bioconductor packages required to use dbSNP
BiocManager::install(version = "3.16")
BiocManager::install("BSgenome")
BiocManager::install("Biostrings")
BiocManager::install("BSgenome.Hsapiens.1000genomes.hs37d5")
BiocManager::install("BSgenome.Hsapiens.NCBI.GRCh38")
BiocManager::install("SNPlocs.Hsapiens.dbSNP155.GRCh37")
BiocManager::install("SNPlocs.Hsapiens.dbSNP155.GRCh38")


```

## Example

see vignette for "tidyGWAS"
```{r}

```


## What does tidyGWAS do?
2.  Remove rows with NAs
3.  Handles duplicates across CHR:POS:REF:ALT or RSID:REF:ALT. Note that this will NOT remove multi-allelics
4.  Can detect and remove indels. Control using `keep_indels`
5.  Update only RSIDs that have been merged into newer ones, using RsMergeArch
6.  Check that RSID follows correct format.
7.  Check for CHR:POS or CHR:POS:REF:ALT in invalid RSIDs, and automatic parsing of this format
8.  Reparation of of RSID for rows with invalid RSID, if format is CHR:POS or CHR:POS:REF:ALT
9.  Validation CHR and POS, automatic conversion of common CHR formats such as 'X', 'M', 'MT' or '23'.
10.  Validation of statistics columns: B, Z, SE, P, N, EAF. See `stats_validation`
11.  Validation of CHR, POS and RSID columns. Identify rows with inconsistent CHR, POS RSID combination.
12.  Repair missing CHR/POS from RSID, or RSID from CHR:POS. 
13.  Remove rows where alleles dont match possible alleles in dbSNP
14.  Remove RSIDs which are not in dbSNP
15.  Add CHR and POS from both GRCh37 and GRCh38
